---
title: "ADD IN MEANINGFUL TITLE HERE"
author: "Group 24: You Yu Fu, Eric Guo, Finn Tran, Lina Jin"
subtitle: "STA304 - Winter 2025 - Assignment 2"
format: pdf
editor: visual
---

```{r, include = FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(broom)
library(knitr)

phone_data <- read_csv("ces_phone.csv")

```

# 1 Introduction

In this section you will briefly describe your report. Explain the importance of the subsequent analysis and prepare the reader for what they will read in the subsequent sections. Provide an overview of the research question. Briefly describe the 2019 Canadian Federal Election Study and its relevance. State the purpose and goals/hypotheses of the report.

# 2 Data

For this project, we will be analyzing the phone survey data set to investigate if education level and age affect voting party preference. To do this, we will be stratifying our data into based on education level, using International Standard Classification of Education (ISCED) levels (Government of Canada, Statistics Canada, 2023) and the Government of Canada 2021 Census Education categories (Government of Canada, Statistics Canada, 2021) as a reference to group our participants based on their response to q61 in the Phone Data Dictionary which asks "What is the highest level of education that you have completed?". By stratifying the sample, we are able to improve the precision of our estimates by accounting for within-group variations of education levels.

| Response | Corresponding Education Level | ISCED Level Definition |
|---------------------------|-------------------|---------------------------|
| No schooling | 0 | Early childhood education |
| Some elementary school | 0 | Early childhood education |
| Completed elementary school | 0 | Early childhood education |
| Some secondary/high school | 0 | Lower secondary education |
| Completed secondary/high school | 1 | High school diploma |
| Some technical, community college | 1 | High school diploma |
| Some university | 1 | High school diploma |
| Completed technical, community college | 2 | Short-cycle tertiary education |
| Bachelor’s degree | 3 | Bachelor or equivalent level |
| Master’s degree | 4 | Master/PhD or equivalent level |
| Professional degree or doctorate | 5 | PhD or equivalent level |

: Our mapping of Phone Data Dictionary Q61 to ISCED_Level and ISCED_Level meaning

We used the above mapping to map respondent answers to ISCED Levels using the ISCED_Level variable. Respondents who picked the option ("Don't know", "Refused to answer" or "Skipped") were removed from the data set, with only 11 out of the over 4000 participants choosing this response. The other variables of interest were cleaned, where we removed all NA values and invalid voter ages (age \< 18). We used q11 to determine surveyor voting choice, where we grouped them into three categories: "Liberal" for those who said they would vote Liberal, "Not Liberal" for those who said they would vote for a party that was not the Liberal party, and "Unknown" for those who skipped the question or chose not to answer.

```{r, include = FALSE}
#NA values in age, most likely to vote for (q11) and education level q(61)
na_count <- phone_data %>%
  summarize(
    q11 = sum(is.na(q11)),
    q61 = sum(is.na(q61)),
    age = sum(is.na(age))
  )

#removing NA values in q11 (81 is a low number)

phone_data <- phone_data %>% filter(!is.na(q11))

#Checking non response answers (-9, -8, -7 for q11 and q61) and valid age

no_response_count <- phone_data %>%
  summarize(
    q11 = sum(q11 < 0),
    q61 = sum(q61 < 0),
    # filtering out ages younger than 18
    age = sum(age < 18)
  )

#removing non response from q61 (11 is a small number)

phone_data <- phone_data %>% filter(q61 > 0)
```

Given that over 25% of phone survey participants chose not to say which party they were going to vote for in the upcoming election, we will generalize this category (q11) into 3 responses, 'Unknown', 'Liberal', and 'Not Liberal'. Removing all these observations could introduce bias into our analysis and decrease the significance of our results.

```{r, include = FALSE}


# Data Cleaning: Convert q61 into ISCED
isced_mapping <- c(
  '1' = 0,  # ISCED 0 -> No school
  '2' = 0,  # ISCED 1 -> Some elem
  '3' = 0,  # ISCED 1 -> Elem
  '4' = 0,  # ISCED 2 -> Some high school
  '5' = 1,  # ISCED 3 -> High school diploma
  '6' = 1,  # ISCED 4 -> Some community college
  '7' = 2,  # ISCED 5 -> Community College
  '8' = 1,  # ISCED 6 -> Some university
  '9' = 3,  # ISCED 6 -> Bachelor
  '10' = 4,  # ISCED 7 -> Master
  '11' = 5  # ISCED 7 -> PHD
)

phone_data <- 
  phone_data %>%
  mutate(isced_level = isced_mapping[as.character(phone_data$q61)],
         voting_choice = case_when(
           phone_data$q11 == 1 ~ 'Liberal',
           phone_data$q11 < 0 ~ 'Unknown',
           phone_data$q11 != 1 ~ 'Non Liberal'
         ))

```

Plotting ISCED_Level (based on q61), and Voting preference (based on q11):

```{r, echo = FALSE}
layout(matrix(c(1, 3, 1, 3, 2, 4, 2, 4), nrow = 4, byrow = TRUE))
#Distribution of Isced Level

barplot(table(phone_data$isced_level),
        main = "Phone Data ISCED Level Distribution",
        xlab = "ISCED Level",
        ylab = "Frequency",
        col = "darkred",
        las = 2) 

boxplot(phone_data$isced_level,
        main = "Phone Data ISCED_Level Boxplot")

#Distribution of voting preference

vote_counts <- phone_data %>% group_by(voting_choice) %>% summarize(count = n())

#Age Distribution

hist(phone_data$age, main = "Phone Data Age Distribution", xlab = "Age", col = "darkblue")

boxplot(phone_data$age,
        main = "Phone Data Age Boxplot")
```

```{r, echo=FALSE}
#Bar plot of liberal voters
ggplot(vote_counts, aes(x = voting_choice, y = count, fill = voting_choice)) +
  geom_bar(stat = "identity") +  # Create bar graph
  geom_text(aes(label = count), vjust = -0.5, color = "black") +  # Add labels
  labs(
    title = "Voting Choice Distribution",
    x = "Voting Choice",
    y = "Count",
    fill = "Voting Choice"
  ) +
  theme_minimal() + 
  scale_fill_manual(values = c("Liberal" = "darkred", "Non Liberal" = "darkblue", "Unknown" = "gray"))
```

```{r, echo=FALSE}
#Summary statistics
print("ISCED LEVEL")
summary(phone_data$isced_level)

print("AGE")
summary(phone_data$age)
```

From the summary statistics and graph, we can see that the survey population is decently well educated, with a mean and median ISCED score of around 2. This means that the average survey respondent has at least completed community college. The ISCED_Level distribution appears to be right skewed, with the interquartile range from 1 to 3 from the box plot and histogram. The age seems slightly right-skewed, with the interquartile range from late 30’s to early 60’s from the box plot, but has a higher range of ages from 18 to late 90’s. Among the web survey respondents, it shows a slight bi-modal distribution on the histogram, with the interquartile range from from late 30’s to early 60’s.

We can make further comparisons on phone survey participants in the 'Unknown' category with those who provided an answer.

```{r, echo = FALSE}
phone_data_unknown <- phone_data %>% filter(voting_choice == 'Unknown')
phone_data_known <- phone_data %>% filter(voting_choice == 'Liberal' | voting_choice == 'Non Liberal')

#Unknown Dist
hist(phone_data_unknown$isced_level, main = "Historgram of ISCED_Level of 'Unknown' Voting Preference", xlab = 'ISCED_Level', col = "black")

#Known Dist
hist(phone_data_known$isced_level, main = "Historgram of ISCED_Level of 'Known' Voting Preference", xlab = 'ISCED_Level', col = "brown")
```

These distributions are very similar, therefore, they could be removed without introducing bias into our data.

```{r, echo=FALSE}
# converting voted_liberal to factor binary (Liberal = 1, Non-Liberal = 0)
phone_data$voting_choice <- factor(phone_data$voting_choice, levels = c("Non Liberal", "Liberal"))

# Ensure that ISCED level and age are numeric
phone_data$isced_level <- as.numeric(phone_data$isced_level)
phone_data$age <- as.numeric(phone_data$age)

phone_data <- phone_data %>%
  mutate(voting_choice = case_when(
    voting_choice == "Liberal" ~ 1,
    TRUE ~ 0
  ))

phone_data <- phone_data %>%
  filter(!is.na(voting_choice))
```

# 3 Methods

To calculate the Confidence Interval for a proportion and stratified sampling, the following formula will be used:

$$
\hat{p} \pm z_{\alpha/2} \sqrt{\sum_{h=1}^{H} W_h^2 \left(1 - \frac{n_h}{N_h} \right) \frac{s_h^2}{n_h}}
$$ Where $\hat{p} = \sum_{h=1}^{H} W_h \hat{p}_h$ and $W_h = \frac{N_h}{N}$

In the formula,

-   $H$ is the number of Education Level strata.

-   $h$ denotes a specific stratum.

-   $\hat{p}_h$ specifies the sample proportion of Liberal voters from the strata $h$.

-   $N_h$ is the population size of the stratum $h$, with numbers from Statistics Canada (Statistics Canada, 2021).

-   $n_h$ is the sample size of the stratum $h$.

-   $s_h^2$ indicates the sample variance of the stratum $h$.

-   $Z_{\alpha/2}$ is the critical value from the standard normal distribution corresponding to a 95% confidence level ($\alpha = 0.05$)

Below is the constructed logistic regression model: $$
\log\left(\frac{1-P(\text{Vote Liberal})}{P(\text{Vote Liberal})}\right)=\beta_0+\beta_1(\text{Education Level})+\beta_2(\text{Age})
$$

-   Education Level, the ordinal variable that the population is stratified by, defining the respondent's education level

-   Age, discrete variable for the respondent's age

-   $P(Vote Liberal)$ describes the probability of the person voting for the Liberal party. The intercept value $\beta_0$ references the expected log-odds of a person with Education Level = 0 and age 0 to vote for the Liberal party. The interpretation is not relevant as a person of age 0 cannot vote.

-   The coefficient $\beta_1$ identifies the expected effect of Education Level in log-odds of voting Liberal compared to the baseline education level (ISCED Level = 0) when Age is kept constant.

-   $\beta_2$ is the expected effect of an unit change in Age on the log-odds of voting Liberal, when Education Level is kept constant.

# 4 Results

Present a table showing the estimated proportion of votes for the selected party along with the 95% confidence interval, and include text describing this table and the key takeaways.

```{r, include = FALSE}

# Here you can run code to calculate your Confidence Intervals.

library(survey)

# Calculate the sample size for each statum
strata_sample_size <- numeric(6)
for (level in 0:5) {
  strata_sample_size[level+1] <- sum(phone_data$isced_level == level)
}

# population size derived from Statistics Canada
phone_data <- phone_data %>% 
  mutate(pop_fpc = case_when(isced_level == 0 ~ 4899580, # No diploma
                             isced_level == 1 ~ 8097295, # High school diploma
                             isced_level == 2 ~ 9252800, # Postsecondary below Bachelor
                             isced_level == 3 ~ 5295360, # Bachelor diploma
                             isced_level == 4 ~ 1740605, # Master's diploma
                             isced_level == 5 ~ 279660)) %>% # Doctorate
  mutate(samp_wt = case_when(isced_level == 0 ~ 4899580/strata_sample_size[1],
                             isced_level == 1 ~ 8097295/strata_sample_size[2],
                             isced_level == 2 ~ 9252800/strata_sample_size[3],
                             isced_level == 3 ~ 5295360/strata_sample_size[4],
                             isced_level == 4 ~ 1740605/strata_sample_size[5],
                             isced_level == 5 ~ 279660/strata_sample_size[6]))

strata_design <- svydesign(id=~1, strata=~isced_level, weights=~samp_wt, fpc=~pop_fpc, data=phone_data)

svymean(~voting_choice, design=strata_design)
confint(svymean(~voting_choice, design=strata_design))
```

In Table 1 I present the confidence intervals of phone surveys:

|   | Proportion of Outcome Variable | 95% Confidence Interval of Outcome Variable |
|-------------------|-----------------------|------------------------------|
| Phone Survey | 0.20914 | (0.1950576, 0.2232315) |

: The proportions and 95% confidence intervals of outcome variable of interest calculated for the Canadian Election Study 2019 phone survey data.

<!-- Provide a table or formula of the estimated logistic regression model, and include text describing this table/formula and the key takeaways. Interpret the estimates from the logistic regression model. Specifically, commenting on how the predictor variables relate to the outcome variable. -->

```{r, echo = FALSE}
logit_model <- svyglm(voting_choice ~ isced_level + age, 
                      design = strata_design, 
                      family = binomial)
```

Below is the estimate regression model:

|             | Estimate ($\beta$) | Standard Error | p-value         |
|-------------|--------------------|----------------|-----------------|
| Intercept   | -2.373643          | 0.163359       | \< 2e-16 \*\*\* |
| ISCED Level | 0.208224           | 0.038918       | 9.28e-08 \*\*\* |
| Age         | 0.012804           | 0.002728       | 2.77e-06 \*\*\* |

Accounting for stratification, -2.373643 is the log-odds intercept for a person with no school on the ISCED scale (ISCED level = 0) and zero years old. This is just a baseline variable, since anyone who is 0 years old cannot vote.

$\beta_1$ is positive and p-value is \< 0.05, meaning that higher education is associated with higher odds of voting Liberal at significance level. The log odds is 0.208224, meaning that for each one-unit increase in education level, the log-odds of voting Liberal increase by around 20.8%.

$\beta_2$ is positive with a log-odds value of 0.012804 and p-value is \< 0.05, meaning that older age is slightly associated with higher log-odds of voting Liberal.

# 5 Discussion

Summarize key findings. Discuss limitations of the analysis (e.g., potential biases, missing variables, survey errors). Provide recommendations for future research or improvements.

# 6 Generative AI Statement

Here is where you can explain your usage of Generative AI tool(s). Be sure to reference any tools with inline citations.

Alternatively, if you did not use Generative AI, please include a brief statement outlining your workflow for completing this assignment.

# 7 Ethics Statement

Explain how you ensured that your analysis is reproducible (e.g., documenting code, using proper statistical methods).

Since the CES 2019 data is publicly available, describe whether or not this the work completed in your report needs Research Ethics Board approval for the report the be made publicly available. Be sure to specifically discuss the privacy of human participants in this study.

\newpage

# 8 Bibliography

1.  Government of Canada, Statistics Canada. “Census Profile, 2021 Census of Populationprofile Table.” *Profile Table, Census Profile, 2021 Census of Population - Canada \[Country\]*, 2 Aug. 2024, www12.statcan.gc.ca/census-recensement/2021/dp-pd/prof/details/page.cfm?LANG=E&GENDERlist=1%2C2%2C3&STATISTIClist=1%2C4&DGUIDlist=2021A000011124&HEADERlist=37&SearchText=Canada.

2.  Government of Canada, Statistics Canada. “International Standard Classification of Education (ISCED).” *Government of Canada, Statistics Canada*, 29 June 2023, www23.statcan.gc.ca/imdb/p3VD.pl?Function=getVD&TVD=1275152.

3.  Grolemund, G. (2014, July 16) *Introduction to R Markdown*. RStudio. <https://rmarkdown.rstudio.com/articles_intro.html>. (Last Accessed: April 4, 1991)

4.  Dekking, F. M., et al. (2005) *A Modern Introduction to Probability and Statistics: Understanding why and how.* Springer Science & Business Media.

5.  Allaire, J.J., et. el. *References: Introduction to R Markdown*. RStudio. <https://rmarkdown.rstudio.com/docs/>. (Last Accessed: April 4, 1991)

\newpage

# 9 Appendix

Any additional notes/derivations that are supplementary to the report can be added in an appendix. This section will not be directly graded, but may be included for completion-sake.
