---
title: "ADD IN MEANINGFUL TITLE HERE"
author: "Group 24: You Yu Fu, Eric Guo, Finn Tran, Lina Jin"
subtitle: "STA304 - Winter 2025 - Assignment 2"
format: pdf
editor: visual
---

```{r, include = FALSE}
library(tidyverse)

phone_data <- read_csv("ces_phone.csv")

```

# 1 Introduction

In this section you will briefly describe your report. Explain the importance of the subsequent analysis and prepare the reader for what they will read in the subsequent sections. Provide an overview of the research question. Briefly describe the 2019 Canadian Federal Election Study and its relevance. State the purpose and goals/hypotheses of the report.

# 2 Data

Briefly introduce the data and key variables of interest. If you do any general data cleaning or data processing you should describe it (in a reproducible manner) here. Identify the stratification variable used. Include at least one plot displaying the distribution of the strata variable. If you do any data cleaning or data processing to the you should describe it (in a reproducible manner) in this section.

```{r, include = FALSE}
#NA values in age, most likely to vote for (q11) and education level q(61)
na_count <- phone_data %>%
  summarize(
    q11 = sum(is.na(q11)),
    q61 = sum(is.na(q61)),
    age = sum(is.na(age))
  )

#removing NA values in q11 (81 is a low number)

phone_data <- phone_data %>% filter(!is.na(q11))

#Checking non response answers (-9, -8, -7 for q11 and q61) and valid age

no_response_count <- phone_data %>%
  summarize(
    q11 = sum(q11 < 0),
    q61 = sum(q61 < 0),
    age = sum(age < 18)
  )

#removing non response from q61 (11 is a small number)

phone_data <- phone_data %>% filter(q61 > 0)
```

Given that over 25% of phone survey participants chose not to say which party they were going to vote for in the upcoming election, I will generalize this category (q11) into 3 responses, 'Unknown', 'Liberal', and 'Not Liberal'. Removing all these observations could introduce bias into our analysis and decrease the significance of our results.

```{r, include = FALSE}


# Data Cleaning: Convert q61 into ISCED
isced_mapping <- c(
  '1' = 0,  # ISCED 0 -> No school
  '2' = 1,  # ISCED 1 -> Some elem
  '3' = 1,  # ISCED 1 -> Elem
  '4' = 2,  # ISCED 2 -> Some high school
  '5' = 3,  # ISCED 3 -> High school diploma
  '6' = 4,  # ISCED 4 -> Some community colege
  '7' = 5,  # ISCED 5 -> Community College
  '8' = 6,  # ISCED 6 -> Some university
  '9' = 6,  # ISCED 6 -> Bachelor
  '10' = 7,  # ISCED 7 -> Master
  '11' = 7  # ISCED 7 -> PHD
)

phone_data <- 
  phone_data %>%
  mutate(isced_level = isced_mapping[as.character(phone_data$q61)],
         voting_choice = case_when(
           phone_data$q11 == 1 ~ 'Liberal',
           phone_data$q11 < 0 ~ 'Unknown',
           phone_data$q11 != 1 ~ 'Non Liberal'
         ))

```

Plotting ISCED_Level (based on q61), and Voting preference (based on q11):

```{r, echo = FALSE}

#Distribution of Isced Level

hist(phone_data$isced_level, main = "Phone Data ISCED_Level Dsitribution", xlab = 'ISCED_Level', col = "darkred")

boxplot(phone_data$isced_level,
        main = "Phone Data ISCED_Level Boxplot")

#Distribution of voting preference

vote_counts <- phone_data %>% group_by(voting_choice) %>% summarize(count = n())

ggplot(vote_counts, aes(x = voting_choice, y = count, fill = voting_choice)) +
  geom_bar(stat = "identity") +  # Create bar graph
  geom_text(aes(label = count), vjust = -0.5, color = "black") +  # Add labels
  labs(
    title = "Voting Choice Distribution",
    x = "Voting Choice",
    y = "Count",
    fill = "Voting Choice"
  ) +
  theme_minimal() + 
  scale_fill_manual(values = c("Liberal" = "darkred", "Non Liberal" = "darkblue", "Unknown" = "gray"))

#Age Distribution

hist(phone_data$age, main = "Phone Data Age Distribution", xlab = "Age", col = "darkblue")

boxplot(phone_data$age,
        main = "Phone Data Age Boxplot")
```

```{r}
#Summary statistics
print("ISCED LEVEL")
summary(phone_data$isced_level)

print("AGE")
summary(phone_data$age)
```

From the summary statistics and graph, we can see that the survey population is decently well educated, with a mean ISCED score of around 5 and a median of 6. This means that the average survey respondent has at least been to community college. The ISCED_Level distribution appears to be left skewed, with the interquartile range from 4 to mostly 6 from the box plot and histogram. The age seems slightly right-skewed, with the interquartile range from late 30’s to early 60’s from the box plot, but has a higher range of ages from 18 to late 90’s. Among the web survey respondents, it shows a slight bi-modal distribution on the histogram, with the interquartile range from from late 30’s to early 60’s(maybe edit, I copy pasted this).

We can make further comparisons on phone survey participants in the 'Unknown' category vs those who provided an answer.

```{r, echo = FALSE}
phone_data_unknown <- phone_data %>% filter(voting_choice == 'Unknown')
phone_data_known <- phone_data %>% filter(voting_choice == 'Liberal' | voting_choice == 'Non Liberal')

#Unknown Dist
hist(phone_data_unknown$isced_level, main = "Historgram of ISCED_Level of 'Unknown' Voting Preference", xlab = 'ISCED_Level', col = "black")

#Known Dist
hist(phone_data_known$isced_level, main = "Historgram of ISCED_Level of 'Known' Voting Preference", xlab = 'ISCED_Level', col = "brown")
```

These distributions are very similar, therefore, they could potentially be removed?

# 3 Methods

Include the formula for calculating the confidence interval for proportions (do not include specific numbers yet) and provide a description of its components. Present the logistic regression model, specifying the independent variables and expected interpretation of coefficients (parameters, not estimates), and describe the model. In this section you are preparing the reader for how to interpret the numbers displayed in the next section (Results).

Be sure to include your CI formula, and a description of the components of the formula.

$$\bar{X} \pm \chi_{\alpha/2} \sqrt{\frac{1+1}{n}}$$

Be sure to include your logistic regression model, and a description of the components of the model.

$$y = mx+ b$$

# 4 Results

Present a table showing the estimated proportion of votes for the selected party along with the 95% confidence interval, and include text describing this table and the key takeaways.

```{r, include = FALSE}

# Here you can run code to calculate your Confidence Intervals.

```

In Table 1 I present both confidence intervals of....

|              | Proportion of Outcome Variable | 95% Confidence Interval of Outcome Variable |
|----------------|-----------------------|---------------------------------|
| Phone Survey | 0.34                           | (lower bound, upper bound)                  |
| Web Survey   | 0.25                           | (lower bound, upper bound)                  |

: The proportions and 95% confidence intervals of outcome variable of interest calculated for both the Canadian Election Study 2019 phone and web survey data.

Provide a table or formula of the estimated logistic regression model, and include text describing this table/formula and the key takeaways. Interpret the estimates from the logistic regression model. Specifically, commenting on how the predictor variables relate to the outcome variable.

Below is the estimate regression model....

$$
\hat{y} = 18x+3
$$

# 5 Discussion

Summarize key findings. Discuss limitations of the analysis (e.g., potential biases, missing variables, survey errors). Provide recommendations for future research or improvements.

# 6 Generative AI Statement

Here is where you can explain your usage of Generative AI tool(s). Be sure to reference any tools with inline citations.

Alternatively, if you did not use Generative AI, please include a brief statement outlining your workflow for completing this assignment.

# 7 Ethics Statement

Explain how you ensured that your analysis is reproducible (e.g., documenting code, using proper statistical methods).

Since the CES 2019 data is publicly available, describe whether or not this the work completed in your report needs Research Ethics Board approval for the report the be made publicly available. Be sure to specifically discuss the privacy of human participants in this study.

\newpage

# 8 Bibliography

1.  Grolemund, G. (2014, July 16) *Introduction to R Markdown*. RStudio. <https://rmarkdown.rstudio.com/articles_intro.html>. (Last Accessed: April 4, 1991)

2.  Dekking, F. M., et al. (2005) *A Modern Introduction to Probability and Statistics: Understanding why and how.* Springer Science & Business Media.

3.  Allaire, J.J., et. el. *References: Introduction to R Markdown*. RStudio. <https://rmarkdown.rstudio.com/docs/>. (Last Accessed: April 4, 1991)

\newpage

# 9 Appendix

Any additional notes/derivations that are supplementary to the report can be added in an appendix. This section will not be directly graded, but may be included for completion-sake.
